Solar wind discontinuities spatial evolution in the outer heliosphere

### 2024-02-06 article.qmd
update

![](figures/juno-hg-loc_sunspot-number)

@@

@@

@@

We will use Liu's [@liu2022a] method to identify discontinuities in the solar wind. This method has better compatibility for the discontinuities with minor field changes, and is more robust to the situation encountered in the outer heliosphere. For each sampling instant $t$, we define three intervals: the pre-interval $[-1,-1/2]\cdot T+t$, the middle interval $[-1/,1/2]\cdot T+t$, and the post-interval $[1/2,1]\cdot T+t$, in which $T$ are time lags. Let time series of the magnetic field data in these three intervals are labeled ${\mathbf B}_-$, ${\mathbf B}_0$, ${\mathbf B}_+$, respectively. Then for an discontinuity, the following three conditions should be satisfied: (1) $\sigma({\mathbf B}_0)>2\max\left(\sigma({\mathbf B}_-, \sigma({\mathbf B}_+)\right)$, (2) $\sigma\left({\mathbf B}_-+{\mathbf B}_+\right)>\sigma({\mathbf B}_-)+\sigma({\mathbf B}_+)$, and (3) $|\Delta {\mathbf B}|>|{\mathbf B}_{bg}|/10$, in which $\sigma$ and ${\mathbf B}_{bg}$ represent the standard deviation and the background magnetic field magnitude, and $\Delta {\mathbf B}={\mathbf B}(t+T/2)-{\mathbf B}(t-T/2)$. The first two conditions guarantee that the field changes of the discontinuity identified are large enough to be distinguished from the stochastic fluctuations on magnetic fields, while the third is a supplementary condition to reduce the uncertainty of recognition. We also will use the minimum or maximum variance analysis (MVA) analysis [@sonnerupMinimumMaximumVariance1998; @sonnerupMagnetopauseStructureAttitude1967] to determine the main (most varying) magnetic field component, $B_l$, and medium variation component, $B_m$. @fig-examples shows several examples of solar wind discontinuities detected by different spacecraft.

@@

## Solar Wind Model

Sadly, JUNO does not provide plasma data during the cruise phase, so to estimate the plasma state we will use MHD model.

We are using [Michigan Solar WInd Model 2D (MSWIM2D)](http://csem.engin.umich.edu/MSWIM2D/), which models the solar wind propagation in 2D using the BATSRUS MHD solver. @keeblerMSWIM2DTwodimensionalOuter2022

Some key points about the model

-   Representing the solar wind in the ecliptic plane from 1 to 75 AU
-   2D MHD model, using the BATSRUS MHD solver
-   Inclusion of neutral hydrogen (important for the outer heliosphere)
-   Inner boundary is filled by time-shifting in situ data from multiple spacecraft

For model validation part, please see [JUNO Model Report](notebooks/20_model.ipynb).

@@

-   We have collected 5 years of solar wind discontinuities from JUNO, aMIS and STEREO.
-   We have developed a pipeline to identify solar wind discontinuities. (Modular, Performant, Scalable)
-   The normalized occurrence rate of IDs drops with the radial distance from the Sun, following $1/r$ law.
-   The thickness of IDs increases with the radial distance from the Sun, but after normalization to ion inertial length, the thickness of IDs decreases.
-   The current intensity of IDs decrease with the radial distance from the Sun, but after normalization to the Alfven current , the current intensity of IDs increases.

@@

# Figures

@@

![Model validation](figures/model/juno_model_validation_full){#fig-modelValidation}

>>>>

![](figures/juno-hg-loc_sunspot-number){#fig-hg-loc-sunspot}

@@

## Dataset

In this study we use datasets of four missions measuring solar wind magnetic field and plasma. Synergistic observations of these missions should advance our understanding of the solar wind discontinuities, their radial distribution and evolution [@velliUnderstandingOriginsHeliosphere2020].

@@

The solar wind plasma state, as evidenced by the sunspot number (referenced in @fig-hg-loc-sunspot and @fig-overview), plays a crucial role in understanding the dynamics of discontinuities. During the initial phase of the JUNO mission, the sunspot number reached its peak, indicating a period of heightened solar activity. However, by the end of the mission's cruise phase, there was a significant decline in solar activity. This variation underscores the importance of calibrating the discontinuity properties in relation to solar activity levels to account for temporal fluctuations.

Furthermore, an analysis of @fig-hg-loc-sunspot reveals the significance of the heliographic longitudinal difference between the Juno mission and Near-Earth missions (such as Wind and ARTEMIS) as well as STEREO-A missions. This longitudinal discrepancy ensures comprehensive coverage of the plasma en route to Juno. Specifically, when there is a substantial longitudinal difference between Juno and Earth, the difference between Juno and STEREO-A tends to be minimal, and vice versa. This relationship facilitates an extensive observation network for the plasma traveling towards Juno.

June measurements do not include plasma characteristics, and to estimate discontinuity spatial scale (thicknesses) we will use solar wind speed obtained from the model of solar wind propagation. The hourly solar wind model data from the Two-Dimensional Outer Heliosphere Solar Wind Modeling (MSWIM2D) [@keebler2022] will be employed to determine the ion bulk velocity $v$ and plasma density $n$ at the location of the Juno mission. Utilizing the BATSRUS MHD solver, this model is capable of simulating the propagation of the solar wind from 1 to 75 astronomical units (AU) in the ecliptic plane, effectively encompassing the region of interest for our study. @fig-model shows comparison of magnetic field magnitudes obtained from MSWIM2D and measured by Juno.

![Model validation](figures/model/juno_model_validation_full){#fig-model}

@@

We will use Liu's [@liuMagneticDiscontinuitiesSolar2022] method to identify discontinuities in the solar wind. This method has better compatibility for the discontinuities with minor field changes, and is more robust to the situation encountered in the outer heliosphere. For each sampling instant $t$, we define three intervals: the pre-interval $[-1,-1/2]\cdot T+t$, the middle interval $[-1/,1/2]\cdot T+t$, and the post-interval $[1/2,1]\cdot T+t$, in which $T$ are time lags. Let time series of the magnetic field data in these three intervals are labeled ${\mathbf B}_-$, ${\mathbf B}_0$, ${\mathbf B}_+$, respectively. Then for an discontinuity, the following three conditions should be satisfied: (1) $\sigma({\mathbf B}_0)>2\max\left(\sigma({\mathbf B}_-, \sigma({\mathbf B}_+)\right)$, (2) $\sigma\left({\mathbf B}_-+{\mathbf B}_+\right)>\sigma({\mathbf B}_-)+\sigma({\mathbf B}_+)$, and (3) $|\Delta {\mathbf B}|>|{\mathbf B}_{bg}|/10$, in which $\sigma$ and ${\mathbf B}_{bg}$ represent the standard deviation and the background magnetic field magnitude, and $\Delta {\mathbf B}={\mathbf B}(t+T/2)-{\mathbf B}(t-T/2)$. The first two conditions guarantee that the field changes of the discontinuity identified are large enough to be distinguished from the stochastic fluctuations on magnetic fields, while the third is a supplementary condition to reduce the uncertainty of recognition. We also will use the minimum or maximum variance analysis (MVA) analysis [@sonnerupMinimumMaximumVariance1998; @sonnerupMagnetopauseStructureAttitude1967] to determine the main (most varying) magnetic field component, $B_l$, and medium variation component, $B_m$. @fig-examples shows several examples of solar wind discontinuities detected by different spacecraft.

@@

# Results

The basic approach of this proposal is to use solar wind measurements at 1AU (STEREO, ARTEMIS, WIND) to compare with Juno and PSP measurements and distinguish effects of solar wind temporal variations and effects of spatial (with the radial distance from the Sun) variations of discontinuity occurrence rate and characteristics. The example of such comparison for the occurrence rate is shown in @fig-rate, where we plot number of discontinuities measured per day by different spacecraft missions (for the same temporal resolution of magnetic field data and the same criteria of discontinuity determination). The radial distance of Juno for 2011-2015 is shown in @fig-model, and the number of discontinuities measured by Juno per day coincides with the discontinuity number measured by STEREO, WIND, and ARTEMIS, when Juno is around $1$ AU. This number (occurrence rate) decreases with distance (with time after $\sim 2013$), as Juno moves from $1$ AU to $5$ AU. We will use the similar comparison for discontinuity characteristics and occurrence rate derived for PSP and Juno.

::: {#fig-rate}
![](figures/ocr/ocr_time_cleaned){width=45%}

The number of discontinuities measured by Juno per day coincides with the discontinuity number measured by STEREO, WIND, and ARTEMIS, when Juno is around $1$ AU. This number (occurrence rate) decreases with distance (with time after $\sim 2013$), as Juno moves from $1$ AU to $5$ AU. We will use the similar comparison for discontinuity characteristics and occurrence rate derived for PSP and Juno. The radial distance of Juno for 2011-2015 is shown in @fig-model.
:::

@@

-   We have collected 5 years of solar wind discontinuities from JUNO, ARTEMIS and STEREO.
-   The normalized occurrence rate of IDs drops with the radial distance from the Sun, following $1/r$ law.
-   The thickness of IDs increases with the radial distance from the Sun, but after normalization to ion inertial length, the thickness of IDs decreases.
-   The current intensity of IDs decrease with the radial distance from the Sun, but after normalization to the Alfven current , the current intensity of IDs increases.

-   We have developed a pipeline to identify solar wind discontinuities. (Modular, Performant, Scalable)

@@

# List of Figures

@@
